Informative Oceanographic Sampling using Excursion Probabilities for Multivariate Random Fields

### 2019-09-25 main_new.tex
Updates from Overleaf

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science. Faced with limited resources to
  understand highly dynamic processes in the water-column, the
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental design for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented,
  for a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

Monitoring the worlds oceans has gained increased importance in the
light of the changing climate and increasing anthropogenic impact. A
central problem for understanding these factors is the lack of
representative data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations on which ocean processes transpire, prompting the need for
effective means of sampling. By \emph{sampling}, we refer to the design
of observational strategies in the spatial domain, where the use of
autonomous platforms can be combined with statistical methods to pursue
measurements with high scientific relevance.

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage, but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have made
strides in advancing the state of sampling and consequently have made
robotics an integral part of ocean observation; our own work
has contributed to this effort.

% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.
Others include \cite{wikle2013modern} focusing on hierarchical
statistical models, \cite{sahu2008space} which studies spatio-temporal
models for sea surface temperature and salinity data, while
\cite{mellucci2018oceanic} looks at the statistical prediction of
features using an underwater glider. Common to these approaches is the
use of information-based methods to infer environmental sampling and overcome common challenges, namely:

\begin{itemize}
\item Decisions on where to sample happen sequentially, i.e.  how
  recently acquired information can be assimilated and used to improve
  statistical models and reduce uncertainty associated with where to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}; this is
  usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization (finding the most resource-efficient design to collect necessary data), and computability (arriving at a
  solution in reasonable time), which has to be considered in
  practice.
\end{itemize}{}

The combination of statistical tools and robotic platforms is a
natural symbiosis which enables information-based sensing. Central to
this is the ability to model spatially-correlated variables and
provide formal measures of uncertainty. By far the most practical
models use Gaussian processes (GPs), which will also be used in our
formulation, in large part because they allow efficient implementation
and evaluation, in turn enabling prioritization of sampling efforts to
be formalized into decision making systems on-board a robotic
platform.

@@

In this work, we build on joint excursion sets (ES) and excursion probabilities (EP) for spatial design of experiments. The goal is to
design sampling strategies for reducing the spatially integrated
Bernoulli variance (IBV) in the ES indicators.  The main contributions
involve: i) extending ES-based uncertainty criteria to multivariate
GPs, allowing the evaluation to include multiple environmental
parameters such as temperature and salinity, ii) developing methods
for informative environmental sensing and autonomous data collection,
and iii) demonstrating the approach in a real world scenario of
sampling a river plume using an AUV.

The remainder of the paper is organized as follows: Section
\ref{sec:bg} provides some background on ocean sampling. Section
\ref{sec:ESEP} defines ESs and EPs, and the design criteria of
IBV. Section \ref{sec:GP_EP} sets up the required statistical modeling
assumptions of the bi-variate GPs for temperature and
salinity. Section \ref{sec:sur} builds on these when deriving the
analytical derivation of the design criteria. Section
\ref{sec:heuristics} extends results to present heuristics for
adaptive sampling. Section \ref{sec:simulations} shows results and
discusses properties of the methods in simulation studies. Section
\ref{sec:case_study} demonstrates the methodology in field work
characterizing a river plume.  Section \ref{sec:concl_disc} contains
summary and a discussion of future work.

@@

Central to understanding the changes taking place in the upper
water-column is the understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography such as fresh water inflows)
mixed with incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem related phenomena such as blooms,
anoxic zones and plumes, all with significant impact both directly and
indirectly on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out that lack of sufficient sampling
resolution in time and space, as a critical issue for characterizing
ocean phenomena. Even today policy makers struggle with limited
observability of severe environmental problems such as the recent (May
2018) outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI}, with losses exceeding 12 million
tons of salmon valued at \$ 80 million (Norwegian Directorate of
Fisheries). This is a case in point in how adequate sampling efforts
could have provided a more accurate assessment and prior
warning. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where the surveying is
guided by a dynamical ocean model and the measurements made onboard
the robot are often used to update the model.

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

Our focus in this work is towards spatial characterization of a
frontal system, and more specifically fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) where cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind and temperature differences, this boundary often gets
distorted. Prior knowledge about the location and evolution of these
features is therefore highly uncertain, making deterministic planning
challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $10 m^2$ -- $5 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit scope to cover only lateral spatial
elements. Moreover, for this case, time-effects can be regarded as
static when acquiring AUV data over a short time window and limited
region. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a range of scientific payloads to survey the
water column. Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000 m with an in-water operation time capacity depending on
survey speed, payload sensors and mission design. By deploying AUVs
together with other resources one can augment information obtained
from either ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator. By instead using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}. The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation. For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling, i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most, has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} which focus largely on reducing
predictive variance or mutual information (entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumption of GP models, and the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest by \cite{Low2009} and \cite{fossuminformation}.
In this paper, with the focus on mapping the river plume, we reward
the designs that improve the classification of water masses, as a
means to characterize the frontal zone.

@@

\section{Excursion sets and Integrated Bernoulli variance}
\label{sec:ESEP}

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater (from the river) and warmer saline waters of the fjord. Excursion sets (ES) and excursion probabilities (EP) are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

We consider a lateral domain, near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes;
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization to more than two variables is given in Section \ref{gen_k}.

We let $t_T$ be a threshold for temperature and $t_S$ a threshold for salinity, the joint ES is then defined by:
\begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_T(\bx) \leq t_T,\xi_S(\bx) \leq t_S\}.
\end{equation}
The excursions could also be defined as larger than a threshold, or
within a boundary of limits for the two variables; calculations and
interpretations will be similar in these cases.

The ES holds random indicator variables, while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is however
challenging if the probability is close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

If prior knowledge provides a strong indication of cold freshwater
(for instance) in a part of the domain, it is easy to classify, and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification skill. We denote temperature measurements by
$y_{T}(\bx_i)$ and salinity measurements by $y_{S}(\bx_i)$, at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements, denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by). 
\end{equation}

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts to locations that are ambiguous; making exploration more effective. As the AUV gathers data throughout its mission, it is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. In this way, we can
take the expectation with respect to the planned measurements $\by$;
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey. A criterion for the selection is:
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}), and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria for situations where the underlying model is based on multivariate GPs.

\section{Gaussian processes and excursion probabilities}
\label{sec:GP_EP}

@@

\subsection{Bivariate Gaussian processes}

>>>>

\begin{abstract}
  Improving and optimizing oceanographic sampling is a crucial task for
  modern marine management and science. Faced with limited resources to
  understand highly dynamic processes in the water-column, the
  combination of statistics and autonomous
  robotics provides opportunities for studying aspects related to
  experimental design for effective characterization of oceanographic
  phenomena. In this work we develop methods for efficient spatial
  sampling applied to the mapping of coastal processes by
  providing informative descriptions of spatial characteristics of ocean
  phenomena. Specifically, we define a design criterion based on
  reducing uncertainty in the joint excursions of random
  fields, and derive analytical expressions for the expected Bernoulli
  variance reduction in the excursion set of spatial temperature and
  salinity variables. We show how these can be applicable in static
  designs and sequential designs, building on myopic and look-ahead
  sampling strategies. We use simulations to study the properties and
  differences between state-of-the-art approaches. Results from field
  deployments with an autonomous underwater vehicle are also presented as part of a case study mapping the boundary of a river plume. The results
  demonstrate the ability of the method to effectively inform and
  execute data-driven environmental sampling.

@@

Monitoring the world's oceans has gained increased importance in 
light of the changing climate and increasing anthropogenic impact. A
central problem for understanding these factors is the lack of
representative data with sufficient resolution. Most of this
\emph{undersampling} can be attributed to the large spatio-temporal
variations on which ocean processes transpire, prompting the need for
effective means of sampling. By \emph{sampling}, we refer to the design
of observational strategies in the spatial domain, where the use of
autonomous platforms can be combined with statistical methods to pursue
measurements with high scientific relevance.

Traditional data collection at sea has typically been based on static
buoys, lagrangian floats, or ship-based methods, with significant
logistical limitations that directly impact coverage and sampling
resolution. Modern methods using satellite remote-sensing provide
large-scale coverage but have limited resolution, are limited to
sensing the surface of the ocean, and are impacted by cloud cover. The
advent of robust mobile robotic platforms \cite{Bellingham07} has
resulted in significant contributions to environmental monitoring and
sampling. In particular, autonomous underwater vehicles (AUVs), have made
strides in advancing the state of sampling and consequently have made
robotics an integral part of ocean observation; our previous work
has contributed to this effort
% using statistical methods in design and data assimilation
\citep{das11b,Graham2013,Das2015,das15,fossuminformation,fossum18b}.
Others working in this space include \cite{wikle2013modern}, focusing on hierarchical
statistical models; \cite{sahu2008space}, studying spatio-temporal
models for sea surface temperature and salinity data; and
\cite{mellucci2018oceanic} looking at the statistical prediction of
features using an underwater glider. Common to these approaches is the
use of information-based methods to infer environmental sampling and overcome common challenges, namely:

\begin{itemize}
\item Fully fledged numerical ocean models cannot be run online on the sensing platforms due to the computational load they require to provide accurate results, hence statistical proxy models of the environment must be used. 
\item Decisions on where to sample happen sequentially, i.e.  
  recently acquired information can be assimilated into and used to improve statistical models and reduce uncertainty associated with where to make measurements.
\item As a priori information about the state of the ocean is very
  uncertain, there is a substantial efficiency gain in reacting to
  information obtained from measurements taken \emph{in-situ}; this is
  usually referred to as the \emph{adaptivity gap} \citep{Krause2008phd}.
\item For such types of sampling problems, the number of choices (i.e.
  locations, trajectories, and candidate designs) makes the problem
  combinatorially large, creating a trade-off between optimization (finding the most resource-efficient design to collect necessary data), and computability (arriving at a
  solution in reasonable time); to engage in successful sampling, this trade-off has to be considered in sample design and practice.
\end{itemize}{}

The combination of statistical tools and robotic platforms is a natural symbiosis which enables information-based sensing. Central to this is the ability to model spatially-correlated variables and provide formal measures of uncertainty. In our formulation we used Gaussian Processes (GPs), in large part because they allow for efficient implementation and evaluation, in turn enabling prioritization of sampling efforts which can be be formalized into decision-making systems on board a robotic platform.

@@

In this work, we build on joint excursion sets (ES) and excursion probabilities (EP) for the spatial design of experiments. The goal is to
design sampling strategies for reducing the spatially-integrated
Bernoulli variance (IBV) in the ES indicators. Our main contributions
include: i) extending ES-based uncertainty criteria to multivariate
GPs, thereby allowing the evaluation to include multiple environmental
parameters such as temperature and salinity; ii) developing methods
for information-based environmental sensing and autonomous data collection;
and iii) demonstrating the approach in the real-world scenario of
sampling a river plume using an AUV.

The remainder of this paper is organized as follows: Section
\ref{sec:bg} provides select background on ocean sampling; Section
\ref{sec:ESEP} defines ESs, EPs, and the design criteria of
IBV; Section \ref{sec:GP_EP} sets up the required statistical modeling
assumptions of the bi-variate GPs for temperature and
salinity; Section \ref{sec:sur} builds on these assumptions when deriving the analytical derivation of the design criteria; Section
\ref{sec:heuristics} extends these results and present heuristics for
adaptive sampling; Section \ref{sec:simulations} shows our results and
discusses properties of the methods in simulation studies. Section
\ref{sec:case_study} demonstrates the methodology used in field work
characterizing a river plume; and Section \ref{sec:concl_disc} contains a summary and a discussion of future work.

@@

Central to understanding the changes taking place in the upper
water-column is an understanding of biophysical interaction driven by
an agglomeration of physical forcing (wind, topography, bathymetry,
tidal influences and local geography, such as fresh water inflows)
and incipient micro-biology driven by planktonic and coastal
anthropogenic input, such as pollution and agricultural runoff. These
often result in a range of ecosystem-related phenomena such as blooms,
anoxic zones and plumes, all of which have significant direct and indirect effects  on society. Walter Munk referred to the
20\textsuperscript{th} century as the \emph{century of undersampling}
\citep{munk2002}, pointing out a lack of sufficient sampling
resolution in time and space as a critical issue for characterizing
ocean phenomena. Even today, policymakers struggle with limited
observability of severe environmental problems such as the recent outbreak of harmful algal blooms in Northern
Norway\footnote{Report from the Institute of Marine
  Research: https://bit.ly/2KZGJfI} (May 2018), with losses exceeding 12 million
tons of salmon valued at \$80 million (Norwegian Directorate of
Fisheries). This is a case in point of how adequate sampling efforts
could have provided a more accurate assessment and prior
warning of changing environmental conditions. These challenges related to undersampling have been driving
ocean sensing practices towards more autonomous, networked, and
distributed use of robotic assets in order to fill the gaps in our
understanding (See Fig. \ref{fig:envir}), where observation is
guided by assimilation of measurements and statistical models.

\begin{figure}[!h] 
  \centering 
  \subfigure[Illustration of an ocean sensing network, where robotic
  assets augment ship-based measurements to increase coverage and
  resolution of phenomena of scientific interest.]{\includegraphics[width =
    0.48\textwidth]{Figures/envir.pdf}\label{fig:envir1}}
  \hfill
  \subfigure[Data-driven sampling under the sense-plan-act paradigm.]{\includegraphics[width =
    0.48\textwidth]{Figures/adaptive_cycle.pdf}\label{fig:sense-plan-act}}
  \caption{Ocean observation is moving away from single-ship sampling
    towards more autonomous and collaborative networked operations in
    order to resolve the numerous bio-geochemical processes and their
    interactions.}
\label{fig:envir}
\end{figure}

Our focus in this work is towards spatial characterization of a
frontal system, more specifically, fronts generated by river
plumes. Fig. \ref{fig:nidelven} shows the survey area (the outlet of
the Nidelva river, Trondheim, Norway) where cold freshwater enters
from the river creating a strong gradient in both temperature and
salinity. Because of the local topography and the Coreolis force
\citep{coriolis1835memoire} the cold fresh water tends to flow near
land to the east. Depending on the river discharge, tidal effects,
wind, and temperature differences, this boundary often gets
distorted. Prior knowledge about the location and evolution of these
features is therefore highly uncertain, making deterministic planning
challenging.

@@

River plumes belong to a class of ocean processes that are local and
at smaller \emph{sub-mesoscale} (from $5 m^2$ -- $10 km^2$) where
spatial lateral variability tends to dominate. Remote sensing and
synthetic ocean models find it challenging to provide detail at this
scale \citep{Lermusiaux:2006} and the principal way to resolve for
fine resolution is by direct observation. At larger \emph{mesoscale}
($>50 km^2$), both three-dimensional space and time dynamics are
important and can shift substantially, while in the case of river
plumes one can often limit the scope to cover only lateral spatial
elements. Moreover, in this case, time-effects can be regarded as
static when acquiring AUV data over a short time window and limited
region. However, the methodological framework can be extended to
higher dimensional situations in a similar manner.

@@

AUVs are used to gather in-situ spatio-temporal data while carrying a range of scientific payloads to survey the
water column. Typically, they operate at $\sim 1-3$ m/s, can reach
depths from 100--6000 m with an in-water operation time capacity dependent on
survey speed, payload sensors and mission design. By deploying AUVs
together with other resources, one can augment information obtained
from ocean models, remote-sensing data, or fixed-location
sensors on buoys. Traditional operations and surveys with AUVs are
usually limited to observations along fixed transects on a spatial
grid, pre-programmed by the human operator. By instead using on-board
algorithms to continuously evaluate, update, and refine future
sampling locations, one can make information gathering adaptive and
contextual to the survey requirements
\citep{das11b,fossuminformation,fossum18b}. The space of sampling
opportunities is still limited to the spatial grid, but now there is
much more flexibility because of adaptation. For instance, an AUV
could reconstruct or modify a survey line based on what temperatures
it is measuring by using embedded methods in decision-theoretic
planning \citep{py10,Rajan12,Rajan12b}.

@@

Adaptive sampling of an evolving frontal feature has been explored in
\cite{fronts11,Zhang2012,Pinto2018,costa19}. These approaches
typically use a reactive-adaptive scheme, whereby exploration does not
rely on a statistical model of the environment, but rather adapts
based on closing the sensing and actuation loop
(Fig. \ref{fig:sense-plan-act}). Myopic sampling—i.e. step-wise
selection of the path (on the graph) which reduces the information
criterion the most—has been used for adaptive surveys
\citep{singh2009efficient,Binney2013} that focus largely on reducing
predictive variance or mutual information (e.g. entropy). Variance and
entropy reduction are independent of the actual data realizations
under the assumptions of GP models; the use of data-driven adaptive
criteria was introduced to include more targeted sampling of regions
of scientific interest in \cite{Low2009} and \cite{fossuminformation}.
In this paper, with a focus on mapping the river plume, we reward
the designs that improve the classification of water masses as a
means to characterize the frontal zone.

@@

\section{Excursion Sets and Integrated Bernoulli Variance}
\label{sec:ESEP}

Our objective is to leverage statistical tools to characterize a river plume, focusing on spatial separation of cold freshwater from the river and warmer saline waters of the fjord. Excursion sets (ES) and excursion probabilities (EP) are useful starting points to measure the ability to classify water masses. The goal of experimental sampling is to improve this characterization, as quantified via changes in the EPs.

We consider a lateral domain near the sea surface, with locations
$\bx \in \mathcal{M} \subset \mathbf{R^2}$. The two variables of
interest, temperature and salinity, are defined as random processes:
$\xi_T(\bx)$ denotes the temperature (in $^o C$) and $\xi_S(\bx)$
denotes the salinity (in $mg/l$) at location $\bx$. The generalization to more than two variables is given in Section \ref{gen_k}.

We let $t_T$ be a threshold for temperature and $t_S$ a threshold for salinity. The joint ES is then defined by:
\begin{equation}\label{ES}
     \mbox{ES} = \{\bx \in \mathcal{M} | \xi_T(\bx) \leq t_T,\xi_S(\bx) \leq t_S\}.
\end{equation}
The excursions could also be defined as larger than a threshold, or
within a boundary of limits for the two variables; calculations and
interpretations will be similar in these cases.

The ES holds random indicator variables while the associated EPs are: 
\begin{equation}\label{eq:prob}
  p(\bx) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S), \hspace{3mm} \bx \in \mathcal{M}.
\end{equation}
If the EPs in Eq. (\ref{eq:prob}) are
close to $1$ or $0$ at a given location, it is easy to classify water mass at this point as either river or ocean; it is, however,
challenging if the probability is close to $0.5$. 
The integrated Bernoulli variance (IBV) \cite{bect2019}, is defined as: 
\begin{equation}\label{measVAR}
    V = \int_{\bx \in \mathcal{M}} p(\bx) \left(1-p(\bx)\right) d\bx.
\end{equation}

If prior knowledge provides a strong indication of cold freshwater
 in a part of the domain, it is easy to classify and
there is likely little value in exploring those locations. Data can be
gathered at selected locations of the domain $\mathcal{M}$, to improve
classification performance. We denote temperature measurements by
$y_{T}(\bx_i)$ and salinity measurements by $y_{S}(\bx_i)$ at a
sampling location $\bx_i \in \mathcal{M}$. Given a set of
measurements denoted $\by$, the conditional EPs are:
\begin{equation}\label{eq:post_ep}
 p(\bx;\by) = P(\xi_T(\bx) \leq t_T, \xi_S(\bx) \leq t_S |\by). 
\end{equation}

Our goal is to construct adaptive sampling strategies based on continuous evaluation of the EPs. As the EPs provide a description of the boundary uncertainty, the strategies can effectively use this description to prioritize sampling efforts at locations that are ambiguous, making exploration more effective. As the AUV gathers data throughout its mission, it is natural to evaluate the expected reduction in the IBV, conditioned on the survey data. In this way, we can
take the expectation with respect to the planned measurements $\by$
\begin{equation}\label{sur}
    V_{\mbox{upd}} = \int_{\bx \in \mathcal{M}} E_{\by} \left\{ p(\bx;\by)\left( 1-p(\bx;\by)\right) \right\} d\bx, 
\end{equation}
where the expectation is over the planned data, and the probability is defined in Eq. (\ref{eq:post_ep}).
Considering a set $\mathcal{J}$ of possible designs for an AUV survey, a criterion for the selection is
\begin{equation}\label{crit}
    j^* = \mbox{argmin}_{j \in \mathcal{J}} V_{\mbox{upd}}(j),
\end{equation}
where the criterion $V_{\mbox{upd}}(j)$ in Eq. (\ref{sur}) is computed
for each of the possible designs.

Recent work on ESs and EPs connected to spatial statistics include
\cite{picheny2010,french2013spatio,bolin2015excursion,french2016credible}.
Our focus is on ES as defined in Eq. (\ref{ES}) and the uncertainty
reduction achieved by sampling as in Eq. (\ref{sur}). In this sense,
our work is similar to \cite{bect2012}, \cite{chevalier2014fast} and
\cite{azzimonti2016quantifying} who describe analytical results for
uncertainty reduction in ESs for univariate processes. A major
contribution of our work is to derive closed-form results for this
design criteria for situations where the underlying model is based on multivariate GPs.

\section{Gaussian Processes and Excursion Probabilities}
\label{sec:GP_EP}

@@

\subsection{Bivariate Gaussian Processes}
